A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-08-28 doc/jcomp2012/jcomp2012.tex
Updated paragraph spacing: Stuart, make sure you only use blank lines in the .tex file where you want new paragraphs to start.

The purpose of this study is to demonstrate an efficient Monte Carlo
solution method for three dimensional (3D), time-dependent, discrete
systems. In this work, we extend our previous work in the following
manner:

\begin{itemize}
\item We develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems.
\item We apply this method to the 1T thermal radiation diffusion
  equation in three dimensions.
\item We compare numerical and performance results of the Monte Carlo
  method against traditional deterministic solution methods.
\end{itemize}

@@

\subsubsection{Direct Method}
\label{sec:direct-method}

@@

\subsubsection{Adjoint Method}
\label{sec:adjoint-method}

@@

\subsection{Monte Carlo Synthetic-Acceleration}
\label{sec:iter-refin-monte}

@@

The MCSA method defined in Eq.~(\ref{eq:MCSA-iteration}) uses the
adjoint method to estimate the error in
Eq.~(\ref{eq:MCSA-residual_solve}) instead of the direct method
outlined in \S~\ref{sec:direct-method}. To demonstrate the
effectiveness of the adjoint method over the direct method within the
context of MCSA, we choose the two-dimensional time-dependent Poisson
equation as a simple model problem:

\begin{equation}
  \frac{\partial \vu}{\partial t} = \nabla^2 \vu\:.
  \label{eq:poisson_equation}
\end{equation}

For all comparisons, a single time step is computed with backwards
Euler time integration. The Laplacian is differenced on a square
Cartesian grid with a second-order five point stencil

\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]\:,
  \label{eq:five_point_stencil}
\end{equation}

and a fourth-order nine point stencil

\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
  \label{eq:nine_point_stencil}
\end{multline}

both assuming a grid size of $\Delta$ in both the $i$ and $j$
directions. For a single time step solution, we then have the
following sparse linear system to be solved with the MCSA method:

\begin{equation}
  \vA \vu^{n+1} = \vu^n\:.
  \label{eq:poisson_eq_lin_sys}
\end{equation}

Both the stencils will be used to vary the size and density of the
sparse linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).

A timing study and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five point and
nine point stencils were used with both the direct and adjoint
solvers. For each case, 50 random walk permutations were computed per
MCSA iteration with a weight cutoff of \sn{1}{-4} for the stochastic
linear solver and a convergence tolerance of \sn{1}{-8} for the MCSA
iterative solver. All computations presented in this section were
completed on a 2.66 GHz Intel Core i7 machine with 8 GB 1067 MHz DDR3
memory. Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for
each case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge
relative to the problem size.

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five point and nine point stencils. A significant
    CPU time speedup is noted with the adjoint method due to the
    reduced number of random walk events.}
  \label{fig:poisson_cpu_time}
\end{figure}

@@

As an additional comparison, the convergence behavior of MCSA can be
analyzed using both solvers to detect any performance benefits. To
assess the convergence properties of MCSA using each solver and
stencil, the infinity norm of the residual computed in
Eq.~(\ref{eq:MCSA-residual}) was collected at each iteration for a
fixed problem size. Fig.~\ref{fig:poisson_convergence} gives the
results of these computations. First, it is worthy to note on the
semilog plot that we are indeed achieving the expected exponential
convergence from MCSA. Second, we note that both the direct and
adjoint methods have approximately equivalent convergence behavior,
leading to CPU time being the dominating factor in method selection.

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of fixed size. Both the adjoint and direct
    solvers are used with the five point and nine point stencils.}
  \label{fig:poisson_convergence}
\end{figure}

@@

The general form of the equilibrium diffusion equation is
\cite{morel_1996}

\begin{equation} 
  (\rho\Cv + 4aT^{3})\pder{T}{t} -
  \nabla\cdot{\Bigl(\frac{4acT^{3}}{3\ros}\Bigr)\grad{T}} = Q\:,
  \label{eq:eq-diff-T}
\end{equation}

where $\Cv\equiv\Cv(\ve{r},T(\ve{r},t))$
$[\text{GJ}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}]$ is the specific
heat capacity of the material, $\rho\equiv\rho(\ve{r})$
$[\text{g}\cdot\text{cm}^{-3}]$ is the density of the material,
$a=0.01372$ $[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{keV}^{-4}]$ is
the radiation constant, $c=29.979$ $[\text{cm}\cdot\text{ns}^{-1}]$ is
the vacuum light speed, and $T\equiv T(\ve{r},t)$ $[\text{keV}]$ is
the temperature that characterizes both the radiation and the
material.  The opacity, $\ros\equiv\ros$(\ve{r},T(\ve{r},t))
$[\text{cm}^{-1}]$, is the Rosseland Mean opacity.  The source,
$Q\equiv Q(\ve{r},t)$
$[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{ns}^{-1}]$, is assumed to
be independent of $T$.

We choose to use the angle-energy integrated radiation intensity,
$\phi\equiv\phi(\ve{r},t)$
$[\text{GJ}\cdot\text{cm}^{-2}\cdot\text{ns}^{-1}]$, as the primary
state variable.  The intensity is defined as:

\begin{equation}
  \label{eq:lte-phi}
  \begin{split}
    \phi(\ve{r},t) &=
    \int_{4\pi}d\bOmega\int_{\nu}d\nu\:\psi(\ve{r},\nu,\hat{\bOmega},t)
    \\ &= \int_{\nu}d\nu\: 4\pi B(\nu,T) \\ &= acT^{4}\:,
  \end{split}
\end{equation}

where $\psi$ is the angular radiation intensity and $\nu$ is the
radiation frequency.  The Planck function (or Planckian) is defined
as,

\begin{equation}
  B(\nu,T)=\frac{2h\nu^{3}}{c^{2}}\bigl(e^{\frac{h\nu}{kT}} -
  1\bigr)^{-1}\:,
\end{equation}

where $h$ is Planck's constant and $k$ is the Boltzmann constant.
Utilizing this definition of $\phi$ we can write

\begin{equation}
  \label{eq:partial-T}
  \partial T = \frac{1}{4acT^{3}}\partial\phi\:.
\end{equation}

Inserting \eqref{eq:partial-T} into \eqref{eq:eq-diff-T} yields

\begin{equation}
  \label{eq:eq-diff}
  \Bigl(\frac{\rho\Cv}{4acT^{3}} + \frac{1}{c}\Bigr)\pder{\phi}{t} -
  \nabla\cdot D\grad{\phi} = Q\:,
\end{equation}

where $D$ is the diffusion coefficient, $D\equiv
D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined as

\begin{equation}
  D = \frac{1}{3\ros}\:.
\end{equation}

Equation~\eqref{eq:eq-diff} is the model equation we will solve using
the Monte Carlo techniques presented in
\S~\ref{sec:monte-carlo-matrix} and \ref{sec:iter-refin-monte}.

@@

Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff})
and lagging the temperature-dependent coefficients at $t^n$ gives

\begin{equation}
  -\nabla\cdot \Dn\grad{\phi} + \sign\phi = \qn\:,
\end{equation}

where we have suppressed the $n+1$ indices on $\phi$ and

\begin{align}
  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\ \qn
  &= \sign\phin + Q^n\:,\\ \phin &= ac(\Tn)^4\:.
\end{align}

Applying Fick's Law,

\begin{equation}
  \ve{F} = -D\grad{\phi}\:,
  \label{eq:fick}
\end{equation}

where $\ve{F}$ is the radiation flux, we define a flux-balance
equation,

\begin{equation}
  \dive{F} + \sign\phi = \qn\:.
  \label{eq:flux-balance}
\end{equation}

Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
Fig.~\ref{fig:cell} gives

\begin{equation}
  \iiint\dive{F}\,d\ve{r} +
  \sign_{ijk}\phi_{ijk}V_{ijk}=\qn_{ijk}V_{ijk}\:,
  \label{eq:flux-balance-vol-integral}
\end{equation}

and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.

\begin{figure}[h!]
  \centerline{ \input{mesh_cell.pdftex_t}}
  \caption{Orthogonal 3D mesh cell.}
  \label{fig:cell}
\end{figure}

We now apply the divergence theorem,

\begin{equation}
  \begin{split}
    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\ &=
    \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
  \end{split}
\end{equation}

to Eq.(\ref{eq:flux-balance-vol-integral}) yielding

\begin{equation}
  \begin{split}
    (F_{i+1/2} - F_{i-1/2})\Dj\Dk + (F_{j+1/2} &- F_{j-1/2})\Di\Dk +
    (F_{k+1/2}\\ &- F_{k-1/2})\Di\Dj + \sign_{ijk}\phi_{ijk}V_{ijk} =
    \qn_{ijk}V_{ijk}\:.
  \end{split}
  \label{eq:flux-balance-difference}
\end{equation}

Using Fick's Law, Eq.~(\ref{eq:fick}), we can write $O(\Delta^2)$
expressions for the face-centered fluxes.  Defining $l\in\{i,j,k\}$,
the face-centered fluxes are

\begin{align}
  F_{l+1/2} &= -D_{l+1/2}\frac{\phi_{l+1} -
    \phi_{l}}{\Delta_{l+1/2}}\:, \label{eq:F_l+1/2}\\ F_{l-1/2} &=
  -D_{l-1/2}\frac{\phi_{l} -
    \phi_{l-1}}{\Delta_{l-1/2}}\:, \label{eq:F_l-1/2}
\end{align}

where

\begin{align}
  \Delta_{l+1/2} &= \frac{\Delta_{l+1} +
    \Delta_{l}}{2}\:,\\ \Delta_{l-1/2} &= \frac{\Delta_{l} +
    \Delta_{l-1}}{2}\:.
\end{align}

In order to be conservative, the flux must be continuous across a
face.  Thus, the face-centered flux evaluated from either side of the
face must be equivalent,

\begin{equation}
  \begin{aligned}
    F_{l+1/2}^{+} &= F_{l+1/2}^{-}\:,\\ 2D_{l+1}\frac{\phi_{l+1} -
      \phi_{l+1/2}}{\Delta_{l+1}} &= 2D_{l}\frac{\phi_{l+1/2} -
      \phi_{l}}{\Delta_l}\:,\\ \frac{1}{3\sigma_{l+1}}\frac{\phi_{l+1}
      - \phi_{l+1/2}}{\Delta_{l+1}} &=
    \frac{1}{3\sigma_{l}}\frac{\phi_{l+1/2} - \phi_{l}}{\Delta_l}\:,
  \end{aligned}
\end{equation}

where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for
clarity.  Solving this system for $\phi_{l+1/2}$ and applying to
either $F_{l+1/2}^{+}$ or $F_{l+1/2}^{-}$ and setting the resulting
expression equal to Eq.~(\ref{eq:F_l+1/2}), one finds the
face-centered diffusion coefficient,

\begin{equation}
  D_{l+1/2} = \frac{2\Delta_{l+1/2}}{3\sigma_l\Delta_l +
    3\sigma_{l+1}\Delta_{l+1}}\:.
  \label{eq:D_l+1/2}
\end{equation}

Following the same procedure at the $l-1/2$ face gives the diffusion
coefficient on the low-side face,

\begin{equation}
  D_{l-1/2} = \frac{2\Delta_{l-1/2}}{3\sigma_l\Delta_l +
    3\sigma_{l-1}\Delta_{l-1}}\:.
  \label{eq:D_l-1/2} 
\end{equation}

Using Eqs.~(\ref{eq:F_l+1/2}), (\ref{eq:F_l-1/2}), (\ref{eq:D_l+1/2}),
and (\ref{eq:D_l-1/2}) in the discrete flux-balance equation,
(\ref{eq:flux-balance-difference}), we obtain

\begin{equation}
  \begin{aligned}
    \sigT\phi_{ijk} -
    \sigm_{i+1}\frac{\Delta_{i+1}}{\Di}\phi_{i+1\,jk} & -
    \sigp_{i-1}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}\\ &-
    \sigm_{j+1}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k} -
    \sigp_{j-1}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}\\ &-
    \sigm_{k+1}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1} -
    \sigp_{k-1}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1} = \qn_{ijk}\:.
  \end{aligned}
  \label{eq:discrete-diffusion}
\end{equation}

Here,

\begin{equation}
  \sigT = \sigp_i + \sigm_i + \sigp_j + \sigm_j + \sigp_k + \sigm_k +
  \sign_{ijk}\:,
\end{equation}

and

\begin{subequations}
  \begin{align}
    \sigp_{i} &= \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di +
      3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\ \sigm_{i} &=
    \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di +
      3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\ \sigp_{j} &=
    \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj +
      3\sigma_{i\,j+1\,k}\Delta_{j+1}}\:,\\ \sigm_{j} &=
    \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj +
      3\sigma_{i\,j-1\,k}\Delta_{j-1}}\:,\\ \sigp_{k} &=
    \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk +
      3\sigma_{ij\,k+1}\Delta_{k+1}}\:,\\ \sigm_{k} &=
    \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk +
      3\sigma_{ij\,k-1}\Delta_{k-1}}\:.
  \end{align}
  \label{eq:sigmas}
\end{subequations}

Also, note that

\begin{align*}
  \sigp_{i-1} &= \frac{1}{\Delta_{i-1}}\frac{2}{3\sigma_{ijk}\Di +
    3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\ \sigm_{i+1} &=
  \frac{1}{\Delta_{i+1}}\frac{2}{3\sigma_{ijk}\Di +
    3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\ \ldots\:.
\end{align*}

Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas})
represent the discretized, 1T diffusion equation on an orthogonal 3D
grid.

@@

Here we derive the equations required to solve
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA
method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
method requires two solutions:

\begin{enumerate}
\item A fixed-point iteration to estimate $\bphi^{l+1/2}$.  This is a
  matrix-vector multiply operation.
\item An adjoint Monte Carlo solve to estimate $\delta\bphi^{l+1/2}$.
\end{enumerate}

For clarity, we rewrite the iteration scheme in
Eq.~(\ref{eq:MCSA-iteration}) using the diffusion operator notation
from Eq.~(\ref{eq:preconditioned-diffusion}),

>>>>

The purpose of this study is to demonstrate an efficient Monte Carlo
solution method for three dimensional (3D), time-dependent, discrete
systems. In this work, we extend our previous work in the following
manner:
\begin{itemize}
\item We develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems.
\item We apply this method to the 1T thermal radiation diffusion
  equation in three dimensions.
\item We compare numerical and performance results of the Monte Carlo
  method against traditional deterministic solution methods.
\end{itemize}

@@

%%---------------------------------------------------------------------------%%
\subsubsection{Direct Method}
\label{sec:direct-method}

@@

%%---------------------------------------------------------------------------%%
\subsubsection{Adjoint Method}
\label{sec:adjoint-method}

@@

%%---------------------------------------------------------------------------%%
\subsection{Monte Carlo Synthetic-Acceleration}
\label{sec:iter-refin-monte}

@@

The MCSA method defined in Eq.~(\ref{eq:MCSA-iteration}) uses the adjoint
method to estimate the error in Eq.~(\ref{eq:MCSA-residual_solve}) instead of
the direct method outlined in \S~\ref{sec:direct-method}. To demonstrate the
effectiveness of the adjoint method over the direct method within the context
of MCSA, we choose the two-dimensional time-dependent Poisson equation as a
simple model problem:
\begin{equation}
  \frac{\partial \vu}{\partial t} = \nabla^2 \vu\:.
  \label{eq:poisson_equation}
\end{equation}
For all comparisons, a single time step is computed with backwards Euler time
integration. The Laplacian is differenced on a square Cartesian grid with a
second-order five point stencil
\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]\:,
  \label{eq:five_point_stencil}
\end{equation}
and a fourth-order nine point stencil 
\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
  \label{eq:nine_point_stencil}
\end{multline}
both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
a single time step solution, we then have the following sparse linear system
to be solved with the MCSA method:
\begin{equation}
  \vA \vu^{n+1} = \vu^n\:.
  \label{eq:poisson_eq_lin_sys}
\end{equation}
Both the stencils will be used to vary the size and density of the sparse
linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).

A timing study and convergence study is used to demonstrate the effectiveness
of the adjoint method as compared to the direct method. To assess both the CPU
time and number of iterations required to converge to a solution, a problem of
constant $\Delta$ was used with varying values of grid size, fixing the
spectral radius of the system at a constant value for each variation. Both the
five point and nine point stencils were used with both the direct and adjoint
solvers. For each case, 50 random walk permutations were computed per MCSA
iteration with a weight cutoff of \sn{1}{-4} for the stochastic linear solver
and a convergence tolerance of \sn{1}{-8} for the MCSA iterative solver. All
computations presented in this section were completed on a 2.66 GHz Intel Core
i7 machine with 8 GB 1067 MHz DDR3 memory. Fig.~\ref{fig:poisson_cpu_time}
gives the CPU time needed for each case to converge in seconds and
Fig.~\ref{fig:poisson_iterations} gives the number of iterations needed for
each case to converge relative to the problem size.
\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five point and nine point stencils. A significant
    CPU time speedup is noted with the adjoint method due to the
    reduced number of random walk events.}
  \label{fig:poisson_cpu_time}
\end{figure}

@@

As an additional comparison, the convergence behavior of MCSA can be analyzed
using both solvers to detect any performance benefits. To assess the
convergence properties of MCSA using each solver and stencil, the infinity
norm of the residual computed in Eq.~(\ref{eq:MCSA-residual}) was collected at
each iteration for a fixed problem size. Fig.~\ref{fig:poisson_convergence}
gives the results of these computations. First, it is worthy to note on the
semilog plot that we are indeed achieving the expected exponential convergence
from MCSA. Second, we note that both the direct and adjoint methods have
approximately equivalent convergence behavior, leading to CPU time being the
dominating factor in method selection.
\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of fixed size. Both the adjoint and direct
    solvers are used with the five point and nine point stencils.}
  \label{fig:poisson_convergence}
\end{figure}

@@

The general form of the equilibrium diffusion equation is \cite{morel_1996}
\begin{equation} 
  (\rho\Cv + 4aT^{3})\pder{T}{t} -
  \nabla\cdot{\Bigl(\frac{4acT^{3}}{3\ros}\Bigr)\grad{T}} = Q\:,
  \label{eq:eq-diff-T}
\end{equation}
where $\Cv\equiv\Cv(\ve{r},T(\ve{r},t))$
$[\text{GJ}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}]$ is the specific heat
capacity of the material, $\rho\equiv\rho(\ve{r})$
$[\text{g}\cdot\text{cm}^{-3}]$ is the density of the material, $a=0.01372$
$[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{keV}^{-4}]$ is the radiation
constant, $c=29.979$ $[\text{cm}\cdot\text{ns}^{-1}]$ is the vacuum light
speed, and $T\equiv T(\ve{r},t)$ $[\text{keV}]$ is the temperature that
characterizes both the radiation and the material.  The opacity,
$\ros\equiv\ros$(\ve{r},T(\ve{r},t)) $[\text{cm}^{-1}]$, is the Rosseland Mean
opacity.  The source, $Q\equiv Q(\ve{r},t)$
$[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{ns}^{-1}]$, is assumed to be
independent of $T$.

We choose to use the angle-energy integrated radiation intensity,
$\phi\equiv\phi(\ve{r},t)$
$[\text{GJ}\cdot\text{cm}^{-2}\cdot\text{ns}^{-1}]$, as the primary state
variable.  The intensity is defined as:
\begin{equation}
  \label{eq:lte-phi}
  \begin{split}
    \phi(\ve{r},t) &=
    \int_{4\pi}d\bOmega\int_{\nu}d\nu\:\psi(\ve{r},\nu,\hat{\bOmega},t)
    \\ &= \int_{\nu}d\nu\: 4\pi B(\nu,T) \\ &= acT^{4}\:,
  \end{split}
\end{equation}
where $\psi$ is the angular radiation intensity and $\nu$ is the radiation
frequency.  The Planck function (or Planckian) is defined as,
\begin{equation}
  B(\nu,T)=\frac{2h\nu^{3}}{c^{2}}\bigl(e^{\frac{h\nu}{kT}} -
  1\bigr)^{-1}\:,
\end{equation}
where $h$ is Planck's constant and $k$ is the Boltzmann constant.  Utilizing
this definition of $\phi$ we can write
\begin{equation}
  \label{eq:partial-T}
  \partial T = \frac{1}{4acT^{3}}\partial\phi\:.
\end{equation}
Inserting \eqref{eq:partial-T} into \eqref{eq:eq-diff-T} yields
\begin{equation}
  \label{eq:eq-diff}
  \Bigl(\frac{\rho\Cv}{4acT^{3}} + \frac{1}{c}\Bigr)\pder{\phi}{t} -
  \nabla\cdot D\grad{\phi} = Q\:,
\end{equation}
where $D$ is the diffusion coefficient, $D\equiv
D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined as
\begin{equation}
  D = \frac{1}{3\ros}\:.
\end{equation}
Equation~\eqref{eq:eq-diff} is the model equation we will solve using the
Monte Carlo techniques presented in \S~\ref{sec:monte-carlo-matrix} and
\ref{sec:iter-refin-monte}.

@@

Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff}) and
lagging the temperature-dependent coefficients at $t^n$ gives
\begin{equation}
  -\nabla\cdot \Dn\grad{\phi} + \sign\phi = \qn\:,
\end{equation}
where we have suppressed the $n+1$ indices on $\phi$ and
\begin{align}
  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\ \qn
  &= \sign\phin + Q^n\:,\\ \phin &= ac(\Tn)^4\:.
\end{align}
Applying Fick's Law,
\begin{equation}
  \ve{F} = -D\grad{\phi}\:,
  \label{eq:fick}
\end{equation}
where $\ve{F}$ is the radiation flux, we define a flux-balance equation,
\begin{equation}
  \dive{F} + \sign\phi = \qn\:.
  \label{eq:flux-balance}
\end{equation}
Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
Fig.~\ref{fig:cell} gives
\begin{equation}
  \iiint\dive{F}\,d\ve{r} +
  \sign_{ijk}\phi_{ijk}V_{ijk}=\qn_{ijk}V_{ijk}\:,
  \label{eq:flux-balance-vol-integral}
\end{equation}
and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.
\begin{figure}[h!]
  \centerline{ \input{mesh_cell.pdftex_t}}
  \caption{Orthogonal 3D mesh cell.}
  \label{fig:cell}
\end{figure}
We now apply the divergence theorem,
\begin{equation}
  \begin{split}
    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\ &=
    \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
  \end{split}
\end{equation}
to Eq.(\ref{eq:flux-balance-vol-integral}) yielding
\begin{equation}
  \begin{split}
    (F_{i+1/2} - F_{i-1/2})\Dj\Dk + (F_{j+1/2} &- F_{j-1/2})\Di\Dk +
    (F_{k+1/2}\\ &- F_{k-1/2})\Di\Dj + \sign_{ijk}\phi_{ijk}V_{ijk} =
    \qn_{ijk}V_{ijk}\:.
  \end{split}
  \label{eq:flux-balance-difference}
\end{equation}

Using Fick's Law, Eq.~(\ref{eq:fick}), we can write $O(\Delta^2)$ expressions
for the face-centered fluxes.  Defining $l\in\{i,j,k\}$, the face-centered
fluxes are
\begin{align}
  F_{l+1/2} &= -D_{l+1/2}\frac{\phi_{l+1} -
    \phi_{l}}{\Delta_{l+1/2}}\:, \label{eq:F_l+1/2}\\ F_{l-1/2} &=
  -D_{l-1/2}\frac{\phi_{l} -
    \phi_{l-1}}{\Delta_{l-1/2}}\:, \label{eq:F_l-1/2}
\end{align}
where
\begin{align}
  \Delta_{l+1/2} &= \frac{\Delta_{l+1} +
    \Delta_{l}}{2}\:,\\ \Delta_{l-1/2} &= \frac{\Delta_{l} +
    \Delta_{l-1}}{2}\:.
\end{align}
In order to be conservative, the flux must be continuous across a face.  Thus,
the face-centered flux evaluated from either side of the face must be
equivalent,
\begin{equation}
  \begin{aligned}
    F_{l+1/2}^{+} &= F_{l+1/2}^{-}\:,\\ 2D_{l+1}\frac{\phi_{l+1} -
      \phi_{l+1/2}}{\Delta_{l+1}} &= 2D_{l}\frac{\phi_{l+1/2} -
      \phi_{l}}{\Delta_l}\:,\\ \frac{1}{3\sigma_{l+1}}\frac{\phi_{l+1}
      - \phi_{l+1/2}}{\Delta_{l+1}} &=
    \frac{1}{3\sigma_{l}}\frac{\phi_{l+1/2} - \phi_{l}}{\Delta_l}\:,
  \end{aligned}
\end{equation}
where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for clarity.
Solving this system for $\phi_{l+1/2}$ and applying to either $F_{l+1/2}^{+}$
or $F_{l+1/2}^{-}$ and setting the resulting expression equal to
Eq.~(\ref{eq:F_l+1/2}), one finds the face-centered diffusion coefficient,
\begin{equation}
  D_{l+1/2} = \frac{2\Delta_{l+1/2}}{3\sigma_l\Delta_l +
    3\sigma_{l+1}\Delta_{l+1}}\:.
  \label{eq:D_l+1/2}
\end{equation}
Following the same procedure at the $l-1/2$ face gives the diffusion
coefficient on the low-side face,
\begin{equation}
  D_{l-1/2} = \frac{2\Delta_{l-1/2}}{3\sigma_l\Delta_l +
    3\sigma_{l-1}\Delta_{l-1}}\:.
  \label{eq:D_l-1/2} 
\end{equation}

Using Eqs.~(\ref{eq:F_l+1/2}), (\ref{eq:F_l-1/2}), (\ref{eq:D_l+1/2}),
and (\ref{eq:D_l-1/2}) in the discrete flux-balance equation,
(\ref{eq:flux-balance-difference}), we obtain
\begin{equation}
  \begin{aligned}
    \sigT\phi_{ijk} -
    \sigm_{i+1}\frac{\Delta_{i+1}}{\Di}\phi_{i+1\,jk} & -
    \sigp_{i-1}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}\\ &-
    \sigm_{j+1}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k} -
    \sigp_{j-1}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}\\ &-
    \sigm_{k+1}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1} -
    \sigp_{k-1}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1} = \qn_{ijk}\:.
  \end{aligned}
  \label{eq:discrete-diffusion}
\end{equation}
Here,
\begin{equation}
  \sigT = \sigp_i + \sigm_i + \sigp_j + \sigm_j + \sigp_k + \sigm_k +
  \sign_{ijk}\:,
\end{equation}
and
\begin{subequations}
  \begin{align}
    \sigp_{i} &= \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di +
      3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\ \sigm_{i} &=
    \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di +
      3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\ \sigp_{j} &=
    \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj +
      3\sigma_{i\,j+1\,k}\Delta_{j+1}}\:,\\ \sigm_{j} &=
    \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj +
      3\sigma_{i\,j-1\,k}\Delta_{j-1}}\:,\\ \sigp_{k} &=
    \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk +
      3\sigma_{ij\,k+1}\Delta_{k+1}}\:,\\ \sigm_{k} &=
    \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk +
      3\sigma_{ij\,k-1}\Delta_{k-1}}\:.
  \end{align}
  \label{eq:sigmas}
\end{subequations}
Also, note that
\begin{align*}
  \sigp_{i-1} &= \frac{1}{\Delta_{i-1}}\frac{2}{3\sigma_{ijk}\Di +
    3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\ \sigm_{i+1} &=
  \frac{1}{\Delta_{i+1}}\frac{2}{3\sigma_{ijk}\Di +
    3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\ \ldots\:.
\end{align*}
Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas}) represent
the discretized, 1T diffusion equation on an orthogonal 3D grid.

@@

Here we derive the equations required to solve
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA method.  As
discussed in \S~\ref{sec:iter-refin-monte}, the MCSA method requires two
solutions:
\begin{enumerate}
\item A fixed-point iteration to estimate $\bphi^{l+1/2}$.  This is a
  matrix-vector multiply operation.
\item An adjoint Monte Carlo solve to estimate $\delta\bphi^{l+1/2}$.
\end{enumerate}
For clarity, we rewrite the iteration scheme in
Eq.~(\ref{eq:MCSA-iteration}) using the diffusion operator notation
from Eq.~(\ref{eq:preconditioned-diffusion}),
\begin{align*}
   \bphi^{l+1/2} = (\vI - \ve{M}^{-1}\ve{D})\bphi^{l} +
  \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\ 
  \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
  \quad & (\text{compute residual})\\ 
  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
  (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\ 
  \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
  (\text{update}\ \bphi)
\end{align*}
As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo method
only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now describe each of these
in turn.
